---
title: "Introduction to consexpressionR"
author: 
  - name: Juliana Costa-Silva
    affiliation:
    - Federal University of Paraná - UFPR and Federal Technological University of Paraná (UTFPR)
    email: julianacostasilvati@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('consexpressionR')`"
vignette: >
  %\VignetteIndexEntry{Introduction to consexpressionR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
library("RefManageR")

bibArticles <- ReadBib("consexpressionR.bib")

bib <- c(
  R = citation(),
  BiocStyle = citation("BiocStyle")[1],
  knitr = citation("knitr")[1],
  RefManageR = citation("RefManageR")[1],
  rmarkdown = citation("rmarkdown")[1],
  sessioninfo = citation("sessioninfo")[1],
  testthat = citation("testthat")[1],
  DESeq2 = citation("DESeq2")[1],
  edgeR = citation("edgeR")[1],
  limma = citation("limma")[1],
  EBSeq = citation("EBSeq")[1],
  NOISeq = citation("NOISeq")[1],
  baySeq = citation("baySeq")[1],
  consexpressionR = citation("consexpressionR")[1]
)

bib <- merge(bib, bibArticles)
```

# Basics

## Background

High-throughput RNA sequencing (RNA-seq) is the standard approach for
quantifying gene expression and identifying genes that are differentially
expressed (DE) between experimental conditions. A large number of methods
have been proposed for this task -- including `r Biocpkg("DESeq2")`,
`r Biocpkg("edgeR")`, `r Biocpkg("limma")`, `r Biocpkg("EBSeq")`,
`r Biocpkg("NOISeq")`, `r Biocpkg("baySeq")`, and `r Biocpkg("KnowSeq")`
-- each based on different statistical assumptions (e.g., negative binomial
generalized linear models, empirical Bayes shrinkage, non-parametric
testing). Because no single method is universally superior across all
experimental designs and noise conditions, the choice of DE method can
materially change the list of genes reported as differentially expressed.

`r Biocpkg("consexpressionR")` addresses this problem by running multiple
independent DE methods on the same count matrix and reporting a
**consensus** list: genes called as differentially expressed by a
user-defined minimum number of methods. The rationale is that genes
detected by several methodologically independent approaches are less
likely to be false positives driven by the assumptions of any single
method.

This consensus strategy was validated against qPCR-confirmed
differentially expressed genes using public datasets (A: available in the Short 
Read Archive (SRA) can be accessed using the code SRA010153 and B: available in 
the Gene Expression Omnibus (GEO) repositorycan accessed using the code 
GSE95077). Results are previouslly reported in 
`r Citep(bib[["Costa-Silva2017"]])` and reviewed in 
`r Citep(bib[["costasilva2025consensus"]])`, showing improved precision compared
to individual methods.

## Install `consexpressionR`

`R` is an open-source statistical environment which can be easily modified to 
enhance its functionality via packages. `r Biocpkg("consexpressionR")` is a `R` 
package available via the [Bioconductor](http://bioconductor.org) repository 
for packages. `R` can be installed on any operating system from 
[CRAN](https://cran.r-project.org/) after which you can install 
`r Biocpkg("consexpressionR")` by using the following commands in your `R` 

session:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("consexpressionR")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

## Required knowledge

`r Biocpkg("consexpressionR")` is built on top of several Bioconductor and
CRAN packages that implement individual differential expression methods:
`r Biocpkg("DESeq2")`, `r Biocpkg("edgeR")`, `r Biocpkg("limma")`,
`r Biocpkg("EBSeq")`, `r Biocpkg("NOISeq")`, and `r Biocpkg("baySeq")` are
required; `r Biocpkg("KnowSeq")` is optional (see
[Optional dependencies](#optional-dependencies) below). Users are
expected to be familiar with the basic structure of a gene-level raw
count matrix (genes as rows, samples as columns) as produced by aligners
such as STAR/HTSeq or pseudo-aligners such as Salmon/kallisto followed by
`r Biocpkg("tximport")`.



<!--
EDIT: confirm whether your package accepts a plain matrix only, or also
a SummarizedExperiment / DESeqDataSet as input. If SummarizedExperiment
is supported (or planned), mention it here and add it as an Import in
DESCRIPTION -- Bioconductor reviewers will likely request this, since it
is the standard container for count data across the project.
-->

If you are asking yourself the question "Where do I start using Bioconductor?" you might be interested in [this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/#.VkOKbq6rRuU).

## Asking for help

As package developers, we try to explain clearly how to use our packages and in which order to use the functions. But `R` and `Bioconductor` have a steep learning curve so it is critical to learn where to ask for help. The blog post quoted above mentions some but we would like to highlight the [Bioconductor support site](https://support.bioconductor.org/) as the main resource for getting help: remember to use the `consexpressionR` tag and check [the older posts](https://support.bioconductor.org/tag/consexpressionR/). Other alternatives are available such as creating GitHub issues and tweeting. However, please note that if you want to receive help you should adhere to the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). It is particularly critical that you provide a small reproducible example and your session information so package developers can track down the source of the error.

## Citing `consexpressionR`

We hope that `r Biocpkg("consexpressionR")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("consexpressionR")
```

# Quick start to using `consexpressionR`

```{r "start", message=FALSE}
library("consexpressionR")
```

## Example dataset

Rather than shipping a static dataset inside the package, this vignette
generates a small synthetic count matrix on the fly using a negative
binomial model -- the same distributional assumption used by most
RNA-seq DE methods. This keeps the package lightweight, keeps the
vignette build fast on the Bioconductor build servers, and lets you
freely change the simulation parameters (number of genes, samples, and
effect size) to explore how the consensus behaves under different
conditions.

```{r "simulate_data"}
set.seed(42)

n_genes <- 200
n_samples_per_group <- 4
prop_de <- 0.10 # proportion of genes that are truly differentially expressed

n_de <- floor(n_genes * prop_de)

## Baseline mean expression per gene (negative binomial mean)
baseline_mu <- rgamma(n_genes, shape = 2, scale = 200)

## Group 2 has a fold-change applied to the first n_de genes
fold_change <- c(rep(4, n_de), rep(1, n_genes - n_de))
mu_group1 <- baseline_mu
mu_group2 <- baseline_mu * fold_change

counts_group1 <- sapply(seq_len(n_samples_per_group), function(i) {
  rnbinom(n_genes, mu = mu_group1, size = 4)
})
counts_group2 <- sapply(seq_len(n_samples_per_group), function(i) {
  rnbinom(n_genes, mu = mu_group2, size = 4)
})

counts <- cbind(counts_group1, counts_group2)
rownames(counts) <- paste0("gene", seq_len(n_genes))
colnames(counts) <- c(
  paste0("control_", seq_len(n_samples_per_group)),
  paste0("treated_", seq_len(n_samples_per_group))
)

sample_info <- c(
  rep("control", n_samples_per_group),
  rep("treated", n_samples_per_group)
)

head(counts)
```

<!--
EDIT: if you would like a fully reproducible "realistic" simulation
instead of this minimal rnbinom() approach, consider using
`r Biocpkg("compcodeR")` or the Polyester package, both of which simulate
RNA-seq count matrices with known ground-truth DE genes and are commonly
used in DE method benchmarking papers. That would also let you show, in
this same vignette, a precision/recall comparison between the consensus
and each individual method -- valuable supporting evidence for reviewers.
-->

## Running individual methods

EDIT: Add 1-2 short examples here calling your individual wrapper
functions directly (e.g., `runDESeq2()`, `runEdgeR()`, `runBaySeq()`),
showing the structure of their output (a `data.frame` with a `DE_<method>`
logical column, as discussed for `runBaySeq()`). For example:

```{r "run_single_method", eval = FALSE}
de_deseq2 <- runDESeq2(counts, sampleInfo = sample_info)
head(de_deseq2)
```

## Running the consensus analysis

EDIT: This is the central example of the vignette and currently missing.
Add the actual call to your main consensus function (its real name --
possibly `consexpression()` or similar -- needs to replace the
placeholder below), explaining the `nMethods` / vote-threshold parameter
clearly, since that is the core contribution of the package:

```{r "run_consensus", eval = FALSE}
result <- consexpression(
  data        = counts,
  sampleInfo  = sample_info,
  minMethods  = 5 # gene must be called DE by at least 5 of 7 methods
)

head(result)
```

EDIT: Show and explain a sample of the *output* table here (which columns
it has -- gene ID, number of methods that agreed, which methods agreed,
consensus logFC, etc.), since that is what most users will want to
parse programmatically afterward.

## Optional dependencies

`r Biocpkg("KnowSeq")` is listed under `Suggests` rather than `Imports`
because of its heavier dependency footprint. If it is not installed,
the corresponding method is skipped with a warning rather than causing
an error, and the consensus is computed using the remaining methods.

```{r "knowseq_check", eval = FALSE}
if (!requireNamespace("KnowSeq", quietly = TRUE)) {
  message("KnowSeq is not installed; it will be skipped in the consensus.")
}
```

EDIT: clarify here whether "5 out of 7" stays "5 out of 6" automatically
when KnowSeq is unavailable, or whether the threshold is fixed regardless
of how many methods actually ran. This is an important behavioral detail
for users and reviewers will likely ask about it directly.

## Interpreting the results

EDIT: Add a short paragraph + (optionally) a plot here. A simple option
that fits naturally into a Bioconductor vignette is an UpSet plot (via
the `r CRANpkg("UpSetR")` package, already in your Suggests) showing the
overlap of DE calls across the 7 methods for this example dataset. That
gives readers an immediate visual sense of *why* consensus filtering
matters, and doubles as a non-trivial example of using your package's
multi-method output.

# Reproducibility

The `r Biocpkg("consexpressionR")` package `r Citep(bib[["consexpressionR"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`
* `r Biocpkg("DESeq2")` `r Citep(bib[["DESeq2"]])`
* `r Biocpkg("edgeR")` `r Citep(bib[["edgeR"]])`
* `r Biocpkg("limma")` `r Citep(bib[["limma"]])`
* `r Biocpkg("EBSeq")` `r Citep(bib[["EBSeq"]])`
* `r Biocpkg("NOISeq")` `r Citep(bib[["NOISeq"]])`
* `r Biocpkg("baySeq")` `r Citep(bib[["baySeq"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
