---
title: "consexpressionR: User guide"
output: 
BiocStyle::html_document:
    toc: FALSE
    number_sections: FALSE
vignette: >
  %\VignetteIndexEntry{A brief Introduction to consexpressionR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# What is it?

<!--http://127.0.0.1:14586/library/limma/doc/intro.html código http://127.0.0.1:14586/library/limma/doc/intro.Rmd -->

consexpressionR is an R package for the analysis of gene expression data, especially the use of consensus analysis, designed experiments and the assessment of differential expression. ConsexpressioR provides the ability to analyse comparisons between many RNA-Seq expression tools simultaneously in a large options designed experiments.

Consensus theory are used to provide stable results. The normalization and background correction functions are provided for RNA-Seq technologies.

<!-- The linear model and differential expression functions apply to a wide variety of gene expression technologies including microarrays (single-channel or two-color), quantitative PCR, RNA-seq or proteomics. -->

```{r setup}
library(consexpressionR)
```
